Towards a Classification of Isolated $j$-invariants

### 2023-07-27 TeX/draft.tex

@@

\newcommand{\davidnote}[1]{\textcolor{Plum}{David: #1}}
\newcommand{\filip}[1]{{\textcolor{cyan}{Filip: [#1]}}}
\newcommand{\abbey}[1]{\textcolor{blue}{Abbey: #1}}
\newcommand{\dld}[1]{\textcolor{Plum}{dld: #1}}
\newcommand{\sachi}[1]{\textcolor{purple}{Sachi: #1}}
\newcommand{\timo}[1]{\textcolor{red}{Timo: #1}}
\newcommand{\travis}[1]{\textcolor{orange}{Travis: #1}}
\newcommand{\abedit}[1]{{\color{blue} #1}}
\newcommand{\fedit}[1]{{\color{cyan} #1}}
% Don't have subsections appear in TOC
%\setcounter{tocdepth}{1}

@@

\section{Introduction}
\label{sec:intro}
\abbey{Updated to account for isolated points.}
The modular curve $X_1(N)$ is an algebraic curve over $\Q$ whose non-cuspidal points parametrize elliptic curves with a distinguished point of order $N$. We are interested in studying isolated points on $X_1(N)$. Roughly speaking these are points of degree $d$ not belonging to an infinite family of degree $d$ points parametrized by a geometric object. For example, if $f: X_1(N) \rightarrow \PP^1$ is a rational map of degree $d$, then $f^{-1}(\PP^1(\Q))$ contains infinitely many closed points of degree $d$ by Hilbert's irreducibility theorem \cite[Chapter 9]{serre97}. We say a degree $d$ point \emph{not} arising from such a map is \textbf{$\PP^1$-isolated}. Other infinite families of degree $d$ points correspond to positive rank abelian subvarieties of the curve's Jacobian; see Section \ref{sec:background} for details. A point which is not thus parametrized is \textbf{AV-isolated}. If a closed point $x\in X_1(N)$ is both $\PP^1$- and AV-isolated, then we say $x$ is \textbf{isolated}. Isolated points can be thought of as those points whose existence is not explained by a geometric construction. The lack of control of isolated points in infinite families of modular curves lies at the heart of several important open problems. See, for example, \cite[Question 1]{MazurNote} and \cite[Theorem 1.3]{BourdonNajman2021}.

One special class of isolated points is \textbf{sporadic} points, which are $x \in X_1(N)$ such that there are only finitely many points on $X_1(N)$ of degree at most $\deg(x)$. Hence a non-cuspidal sporadic point corresponds to an elliptic curve with a point of order $N$ in ``usually low degree." Elliptic curves with complex multiplication (CM) provide many natural examples of sporadic points, since the extra endomorphisms of a CM elliptic curve constrain the size of the image of the associated Galois representation. Indeed, as shown in \cite[Theorem 8.2]{CGPS2022}, there exist sporadic CM points on $X_1(N)$ for all $N\geq 721$.

Non-CM isolated points on $X_1(N)$ remain much more mysterious. One recent line of investigation has focused on the class of isolated points associated to non-CM elliptic curves with $j$-invariant in $\Q$. To date, there are only three known examples of such curves, up to isomorphism over $\overline{\Q}$:
\begin{itemize}
\item The elliptic curve with $j$-invariant $-140625/8$ corresponds to two points of degree 3 on $X_1(21)$. Since there are only finitely many points of degree at most 3 on this modular curve, these are sporadic (and hence isolated) points. This example was first discovered by Najman \cite{najman16}. In fact, this is the unique elliptic curve giving a sporadic point of degree at most 3 on \emph{any} modular curve of the form $X_1(N)$, as shown in recent work of Derickx, Etropolski, van Hoeij, Morrow, and Zureick-Brown \cite{DEvHMZB2021}.
\item The elliptic curve with $j$-invariant $-9317$ gives three degree 6 points on $X_1(37)$, as in work of van Hoeij \cite{vanHoeij}. Since 6 is less than half the $\Q$-gonality of $X_1(37)$, as computed in \cite{DerickxVanHoeij2014}, the point is necessarily sporadic by work of Frey \cite{frey}.
\item The elliptic curve with $j$-invariant $351/4$ gives an isolated point of degree 9 on $X_1(28)$; see \cite[Theorem 2]{OddDeg}. There are infinitely many points on $X_1(28)$ of degree 6, as shown in \cite{DerickxVanHoeij2014}, so this point is isolated but not sporadic.
\end{itemize}

@@

In \cite{BELOV}, they ask whether one can explicitly determine the set of isolated $j$-invariants in $\Q$. This is the motivation for the present work. We develop an algorithm which can be used to determine whether a given non-CM $j$-invariant $j\in \Q$ is isolated. Starting with the adelic image of the adelic Galois representation associated $E/\Q$ with $j(E)=j(x)$, as computed by Zywina \cite{ZywinaAlgorithm}, we apply various filters inspired by \cite{BELOV} to determine whether there exists an isolated point on $x \in X_1(N)$ with $j(x)=j$.

\begin{algorithm}[h!]
\caption{Main Algorithm}
\KwIn{A non-CM $j$-invariant $j \in \Q$.}
\KwOut{A list $[j,\{\langle a_1,d_1\rangle, \dots, \langle a_k,d_k\rangle \}]$ such that any isolated point $x \in X_1(N)$ with $j(x)=j$ maps down under the natural projection map to an isolated point of degree $d_i$ on $X_1(a_i)$ for some $1 \leq i \leq k$.}
\end{algorithm}

\noindent In particular, if the output of Algorithm \ref{alg:mainalgorithm} is the empty set, then $j$ is not an isolated $j$-invariant. We give an overview of the main algorithm in Section \ref{alg:mainalgorithm}, with key sub-algorithms described in detail in Sections \ref{sec:madicrep} and \ref{sec:primitivedeg}. We address the validity of Algorithm \ref{alg:mainalgorithm} in Section \ref{sec:validity}.

Running Algorithm \ref{alg:mainalgorithm} on all elliptic curves currently in the $L$-functions and Modular Forms Database (LMFDB) \cite{LMFDB} gives a list of 95 $j$-invariants which require further consideration; see the Appendix. After applying certain additional filters to address special cases, as discussed in Section~\ref{sec:remainingfilters}, we arrive at the following result.

\begin{theorem}\label{LMFDBoutputThm}
Let $x=[E,P]\in X_1(N)$ be a non-CM isolated point with $j(E) \in \Q$.
Fix an equation for $E/\Q$ and let $N_E$ denote its conductor.
Suppose that one of the following holds:
\begin{itemize}
    \item$N_E \leq 500{,}000$,
    \item $N_E$ is only divisible by primes $p \leq 7$, or
    \item $N_E=p \leq 200{,}000{,}000$ for some prime number $p$.
\end{itemize}
Then $j(E) \in \{-140625/8,-9317,351/4, -162677523113838677\}$. Moreover, each of these $j$-invariants corresponds to a $\PP^1$-isolated point on $X_1(21)$, $X_1(37)$, $X_1(28)$, and $X_1(37)$, respectively.
\end{theorem}

Based on these results, we conjecture that the four non-CM isolated $j$-invariants identified above are in fact the \emph{only} $j$-invariants in $\Q$ associated to non-CM elliptic curves which give rise to $\PP^1$-isolated points.

\begin{conjecture}
If $x\in X_1(N)$ is a $\PP^1$-isolated point with $j(x) \in \Q$, then $j(x)=-140625/8$,$-9317$, $351/4$, $-162677523113838677$, or one of the 13 CM $j$-invariants in $\Q$.
\end{conjecture}

\noindent Since any CM elliptic curve is known to produce sporadic points on infinitely many modular curves of the form $X_1(N)$ by \cite[Theorem 7.1]{BELOV}, it follows conversely that every $j$-invariant in this set is isolated.

It is worth noting that the sporadic points on $X_1(21)$ associated to $j=-140625/8$ and the sporadic points on $X_1(37)$ associated to $j=-9317$ lie above exceptional rational points on $X_0(21)$ and $X_0(37)$, respectively. We say a point $x \in X_0(N)(\Q)$ is \textbf{exceptional} if $X_0(N)(\Q)$ is finite and $x$ corresponds to a non-CM elliptic curve over $\Q$. Thus it is natural to ask whether there exist other sporadic points on $X_1(N')$ lying above such exceptional rational points. Running Algorithm~\ref{alg:mainalgorithm} on the 14 $j$-invariants corresponding to exceptional rational points on $X_0(N)$ (see, for example, \cite[Table 4]{LRAnn}) shows that there are no additional sporadic $j$-invariants in this set.
\begin{theorem}
Let $E$ be an elliptic curve corresponding to an exceptional rational point on $X_0(N)$. If there exists a sporadic point $x \in X_1(N')$ with $j(x)=j(E)$, then $j(E) =-140625/8$ or $-9317$.
\end{theorem}

@@

@@

\subsection{Modular Curves}
For any subgroup $H \leq \GL_2(\Z/N\Z)$, we define the modular curve $X_H$ to be the coarse space of the stack $\mathcal{M}_H$, as defined in Deligne--Rapoport \cite{DR}. The curve $X_H$ is a scheme over $\text{Spec }\Z[1/N]$ and parametrises generalized elliptic curves with $H$-level structure. \abedit{In particular, its $k$-rational points roughly classify elliptic curves over $k$ whose mod $N$ image is contained in $H$}; see, for example, \cite[$\S2.3$]{RSZB2022} for details. If we take
\[
B_1(N) =\left\{ \begin{pmatrix}
1 & * \\
0 & *
\end{pmatrix} \right \} \leq \GL_2(\Z/N\Z),
\]
then $X_{B_1(N)}=X_1(N)$, the modular curve whose noncuspidal points parametrize  elliptic curves with a distinguished point of order $N$. There is an analytic isomorphism between $X_1(N)(\mathbb{C})$ and the Riemann surface constructed as a quotient of the extended upper-half plane by the congruence subgroup
\[
\Gamma_1(N) \coloneqq\left\{ \begin{pmatrix}
a & b \\
c & d 
\end{pmatrix} \in \SL_2(\Z) \, : \, c \equiv 0 \pmod{N} \text{ and } a \equiv d \equiv 1 \pmod{N} \right \}, 
\]
with matrices acting via linear fractional transformations. If $N \geq 4$, then $\mathcal{M}_{B_1(N)}$ is its own coarse moduli space, and so noncuspidal $k$-rational points of $X_1(N)$ classify pairs $(E,P)$, where $E/k$ is an elliptic curve and $P\in E(k)$, up to $k$-isomorphism. 
%\abbey{I decided not to define the modular curve by its moduli problem as in \cite[$\S2$]{RouseDZB}, since the construction yields a stack in general... so it seems confusing to call it a modular *curve*. I've opted to go with the conventions of \cite[$\S2.3$]{RSZB2022}.}

@@

\subsection{Closed Points on Modular Curves}\abbey{I updated the first 2 paragraphs of this section.} To discuss isolated points on modular curves, we must consider closed points on $X_1(N)$, viewed always as a scheme over $\Q$. Let $k$ be a field with an embedding of $k$ into $\overline \Q$. Given an elliptic curve $E/k$ with $P \in E(k)$ of order $N$, the pair $(E,P)$ induces a $k$-valued point on $X_1(N)$ via the moduli interpretation described above. We denote this $k$-valued point by $(E,P)_k$, and by definition it corresponds to a morphism of $\Q$-schemes $f:\text{Spec } k \rightarrow X_1(N)$. The map $f$ sends the unique point of $\text{Spec } k$ to a point $x \in X$, and we call $x$ the \textbf{closed point} associated to $(E,P)$.\footnote{Note $x$ is indeed closed since $\Q(x)/\Q$ is finite; see, for example, \cite[Exercise 5.9, p.76] {Liu2002}.} We define the \textbf{degree} of $x$ to be the degree of the residue field $\Q(x)$ over $\Q$. Since there are many scheme-valued points which induce the same closed point, it is sometimes preferable to consider Galois orbits of points in $X_1(N)(\overline{\Q})$, which are in bijection with the set of closed points. Thus one could alternatively define the closed point associated to $(E,P)$ as the $\Gal_\Q$-orbit of $(E,P)_{\overline{\Q}}$. \sachi{I think the difference between $k$-valued points and $k$-rational points is confusing, since the names are so similar. It would be nice to explicitly point out what is meant by $k$-rationality somewhere (in constrast to $k$-valued).} \abbey{They are all the same thing for a scheme over a field. I used the term $k$-valued b/c it's the term used in Qing Liu's book \cite{Liu2002}, but it matches what Poonen calls a $k$-rational point or a $k$-point of a scheme over a field in \emph{Rational points on varieties}, for example. The definition is actually already above --- it's a morphism of $\Q$-schemes $f:\text{Spec} k \rightarrow X_1(N)$. If you say $k$-rational point, I imagine most people just think of points with coordinates in $k$, rather than this scheme-theoretic definition. These two notions coincide, but it's something to prove, for example, as in Prop. 2.18 in Section 3.2.3 of \cite{Liu2002}.}

@@

\begin{remark}\label{ResidueFieldRmk} In particular, if $E/\Q(j(E))$ does not have complex multiplication and $P=(x_0,y_0)$, then one can take $\Q(x)\cong\Q(j(E),x_0)$. See \cite[p. 107]{shimura}.
\end{remark}
\begin{example}
Let $E_1:y^2+xy+y=x^3-x^2-3x+3$ and $P_1=(-1,2)$ a point of order 7. Then $(E_1,P_1)$ gives a $\Q$-valued point on $X_1(7)/\Q$ and also a closed point $x\in X_1(7)$ of degree 1. On the other hand, let $E_2:y^2=x^3-43x-166$ and $P_2=(5,i\sqrt{256})$ a point of order 7. Then $(E_2,P_2)$ gives a $\Q(i)$-valued point on $X_1(7)/\Q$ and also a closed point $x\in X_1(7)$ of degree 1 by \abedit{Remark \ref{ResidueFieldRmk}} \sachi{(and therefore a $\Q$-rational point)} \abbey{We want to think of it as \emph{inducing} a $\Q$-rational point---as explained in the paragraph below, or in Remark 7 above---rather than $(E_2,P_2)$ itself being $\Q$-rational. The non-cuspidal $\Q$-rational points of $X_1(7)(\Q)$ correspond to $\Q$-isomorphism classes of elliptic curves $E/\Q$ with a point $P\in E(\Q)$ of order 7, where this comes from the curve being a fine moduli space. So the representative $E_2,P_2$ does not show up here, since $P_2$ is not in $\Q$.}. In fact, both $(E_1,P_1)$ and $(E_2,P_2)$ induce the same geometric point on $X_1(7)$ since $(E_1,P_1)_{\overline{\Q}}=(E_2,P_2)_{\overline{\Q}}$.

@@

To determine its degree, let $\Gamma$ and $\Gamma'$ be the intersection with $\SL_2(\Z)$ of the inverse image of $G$ and $G'$ in $\GL_2(\Z)$. Over $\bC$, the morphism $f$ is the quotient map $\Gamma \backslash \mathcal{H} \rightarrow \Gamma' \backslash \mathcal{H}$. Since the kernel of the action of $\SL_2(\Z)$ on $\mathcal H$ is $\pm I$, the degree of $f$ is as claimed.
    \end{proof}
Let $a$ and $b$ be positive integers. Taking $G=\pi^{-1}(B_1(ab))$ and $G'=\pi^{-1}(B_1(a))$ gives the following corollary, which under the moduli interpretation corresponds to sending $(E,P)$ to $(E,bP)$.    \begin{corollary}
                For positive integers $a$ and $b$, the natural $\Q$-rational map $f\colon X_1(ab) \rightarrow X_1(a)$ has
                \[
                    \deg(f)=
                    c_{f}\cdot b^2 \prod_{p \mid b,\, p \nmid a}
                    \left(1-\frac{1}{p^2}\right),
                \]
                where $c_{f}=1/2$ if $a \leq 2$ and $ab>2$ and $c_{f}=1$ otherwise. 
    \end{corollary}

>>>>

%for tables
\usepackage{tabularx}
\usepackage{booktabs}

@@

\newcommand{\davidnote}[1]{\textcolor{Plum}{David: #1}}
\newcommand{\filip}[1]{{\textcolor{cyan}{Filip: [#1]}}}
\newcommand{\abbey}[1]{\textcolor{blue}{Abbey: #1}}
\newcommand{\dld}[1]{\textcolor{Plum}{dld: #1}}
\newcommand{\sachi}[1]{\textcolor{purple}{Sachi: #1}}
\newcommand{\timo}[1]{\textcolor{red}{Timo: #1}}
\newcommand{\travis}[1]{\textcolor{orange}{Travis: #1}}
\newcommand{\abedit}[1]{{\color{blue} #1}}
\newcommand{\fedit}[1]{{\color{cyan} #1}}
\newcommand{\sedit}[1]{{\color{purple} #1}}

% Don't have subsections appear in TOC
%\setcounter{tocdepth}{1}

@@

\section{Introduction}
\label{sec:intro}

The modular curve $X_1(N)$ is an algebraic curve over $\Q$ whose non-cuspidal points parametrize elliptic curves with a distinguished point of order $N$. We are interested in studying isolated points on $X_1(N)$. Roughly speaking these are points of degree $d$ not belonging to an infinite family of degree $d$ points parametrized by a geometric object. For example, if $f: X_1(N) \rightarrow \PP^1$ is a rational map of degree $d$, then $f^{-1}(\PP^1(\Q))$ contains infinitely many closed points of degree $d$ by Hilbert's irreducibility theorem \cite[Chapter~9]{serre97}. We say a degree $d$ point \emph{not} arising from such a map is \textbf{$\PP^1$-isolated}. Other infinite families of degree $d$ points correspond to positive rank abelian subvarieties of the curve's Jacobian; see Section \ref{sec:background} for details. A point which is not thus parametrized is \textbf{AV-isolated}. If a closed point $x\in X_1(N)$ is both $\PP^1$- and AV-isolated, then we say $x$ is \textbf{isolated}. Isolated points can be thought of as those points whose existence is not explained by a geometric construction. The lack of control of isolated points in infinite families of modular curves lies at the heart of several important open problems. See, for example, \cite[Question~1]{MazurNote} and \cite[Theorem~1.3]{BourdonNajman2021}.

One special class of isolated points is \textbf{sporadic} points, which are $x \in X_1(N)$ such that there are only finitely many points on $X_1(N)$ of degree at most $\deg(x)$. Hence a non-cuspidal sporadic point corresponds to an elliptic curve with a point of order $N$ in ``usually low degree." Elliptic curves with complex multiplication (CM) provide many natural examples of sporadic points, since the extra endomorphisms of a CM elliptic curve constrain the size of the image of the associated Galois representation. Indeed, as shown in \cite[Theorem~8.2]{CGPS2022}, there exist sporadic CM points on $X_1(N)$ for all $N\geq 721$.

Non-CM isolated points on $X_1(N)$ remain much more mysterious. One recent line of investigation has focused on the class of isolated points associated to non-CM elliptic curves with $j$-invariant in $\Q$. To date, there are only three known examples of such curves, up to isomorphism over $\overline{\Q}$:
\begin{itemize}
    \item The elliptic curve with $j$-invariant $-140625/8$ corresponds to two points of degree 3 on $X_1(21)$. Since there are only finitely many points of degree at most 3 on this modular curve, these are sporadic (and hence isolated) points. This example was first discovered by Najman \cite{najman16}. In fact, this is the unique elliptic curve giving a sporadic point of degree at most 3 on \emph{any} modular curve of the form $X_1(N)$, as shown in recent work of Derickx, Etropolski, van Hoeij, Morrow, and Zureick-Brown \cite{DEvHMZB2021}.

    \item The elliptic curve with $j$-invariant $-9317$ gives three degree 6 points on $X_1(37)$, as in work of van Hoeij \cite{vanHoeij}. Since 6 is less than half the $\Q$-gonality of $X_1(37)$, as computed in \cite{DerickxVanHoeij2014}, the point is necessarily sporadic by work of Frey \cite{frey}.

    \item The elliptic curve with $j$-invariant $351/4$ gives an isolated point of degree 9 on $X_1(28)$; see \cite[Theorem 2]{OddDeg}. There are infinitely many points on $X_1(28)$ of degree 6, as shown in \cite{DerickxVanHoeij2014}, so this point is isolated but not sporadic.
\end{itemize}

@@

In \cite{BELOV}, they ask whether one can explicitly determine the set of isolated $j$-invariants in $\Q$. This is the motivation for the present work. We develop an algorithm which can be used to determine whether a given non-CM $j$-invariant $j\in \Q$ is isolated. Starting with the adelic image of the adelic Galois representation associated $E/\Q$ with $j(E)=j(x)$, as computed by Zywina \cite{ZywinaAlgorithm}, we apply various filters to determine whether there exists an isolated point on $x \in X_1(N)$ with $j(x)=j$.

\begin{algorithm}[h!]
\caption{Main Algorithm}
    \KwIn{A non-CM $j$-invariant $j \in \Q$.}
    \KwOut{A \abedit{finite} list $[j,\{\langle a_1,d_1\rangle, \dots, \langle a_k,d_k\rangle \}]$ such that any isolated point $x \in X_1(N)$ with $j(x)=j$ maps down under the natural projection map to an isolated point of degree $d_i$ on $X_1(a_i)$ for some $1 \leq i \leq k$.}
\end{algorithm}

\noindent In particular, if the output of Algorithm \ref{alg:mainalgorithm} is the empty set, then $j$ is not an isolated $j$-invariant.

One of the key filters we use is based on \cite[Theorem 4.3]{BELOV}.

\begin{theorem}[\cite{BELOV}]
Let $f: C \rightarrow D$ be a finite map of curves, let $x \in C$ be a closed point, and let $y=f(x)\in D$. Assume that $\deg(x)=\deg(y)\cdot \deg(f)$.
\begin{enumerate}
\item If $x$ is $\PP^1$-isolated, then $y$ is $\PP^1$-isolated.
\item If $x$ is AV-isolated, then $y$ is AV-isolated.
\item $x$ is sporadic, then $y$ is sporadic.
\end{enumerate}
\end{theorem}

\noindent Since the image of the adelic Galois representation of a non-CM elliptic curve $E/\Q$ has finite level, this theorem reduces the problem of determining whether $j(E)$ is isolated to the problem of determining whether $E$ gives rise to an isolated point on a finite number of modular curves. While a sufficient finite list can be deduced directly from the adelic level as found by Zywina \cite{ZywinaAlgorithm}, our output works to reduce this list to its minimal size. This inspires the notion of \textbf{primitive points} on modular curves, as studied in Section \ref{sec:PrimitivePoints}. In particular, $j(E)$ is isolated if and only if there exists an isolated primitive point corresponding to $E$. Moreover, any point $x\in X_1(N)$ with $j(x)=j(E)$ corresponds to a \emph{unique} primitive point (Proposition \ref{Prop:single_sink}).

The second part of Algorithm \ref{alg:mainalgorithm} works to show the primitive points corresponding to $E$ are \emph{not} isolated. For example, if the Riemann-Roch space associated to $x \in X_1(N)$ with $j(x)=j(E)$ has dimension at least 2, then $x$ is not $\PP^1$-isolated (and hence not isolated). In other cases, we can show $x \in X_1(N)$ is not isolated by applying the following result.

\begin{proposition} \label{Prop:Genus0Intro}
Let $E/\Q$ be an elliptic curve, and let $H \leq \GL_2(\Z/N\Z)$ be the image of the mod $N$ Galois representation of $E$, after some choice of basis. If the modular curve $X_H$ has genus 0, then there are no isolated points on $X_1(N)$ associated to $E$.
\end{proposition}

\noindent In particular, this result implies any elliptic curve with adelic image of genus 0 does not give rise to any isolated points on $X_1(N)$, even as $N$ ranges over all positive integers. However, even when the adelic image of $E$ has positive genus, it can still be that $\im \rho_{E,N}$ has genus 0 for all $N$ corresponding to primitive points. This makes Proposition \ref{Prop:Genus0Intro} more broadly applicable.

Running Algorithm \ref{alg:mainalgorithm} on all elliptic curves currently in the $L$-functions and Modular Forms Database (LMFDB) \cite{LMFDB} gives a list of 95 $j$-invariants which require further consideration; see the Appendix. \abbey{I think we should build Prop. \ref{Prop:Genus0Intro} into the algorithm. Would need to update to mention only Table 1 instead.} After applying certain additional filters to address special cases, as discussed in Section~\ref{sec:remainingfilters}, we arrive at the following result.

\begin{theorem}\label{LMFDBoutputThm}
Let $x=[E,P]\in X_1(N)$ be a non-CM isolated point with $j(E) \in \Q$.
Fix an equation for $E/\Q$ and let $N_E$ denote its conductor.
Suppose that one of the following holds:
\begin{itemize}
    \item $N_E \leq 500{,}000$,
    \item $N_E$ is only divisible by primes $p \leq 7$, or
    \item $N_E=p \leq 200{,}000{,}000$ for some prime number $p$.
\end{itemize}
Then $j(E) \in \{-140625/8,-9317,351/4, -162677523113838677\}$. Moreover, each one of these $j$-invariants corresponds to a $\PP^1$-isolated point on $X_1(21)$, $X_1(37)$, $X_1(28)$, and $X_1(37)$, respectively.
\end{theorem}

Based on these results, we conjecture that the four non-CM isolated $j$-invariants identified above are in fact the \emph{only} $j$-invariants in $\Q$ associated to non-CM elliptic curves which give rise to $\PP^1$-isolated points. Since any CM elliptic curve is known to produce sporadic points on infinitely many modular curves of the form $X_1(N)$ by \cite[Theorem 7.1]{BELOV}, it follows conversely that every $j$-invariant in this set is isolated.

\begin{conjecture}\label{conj}
If $x\in X_1(N)$ is a $\PP^1$-isolated point with $j(x) \in \Q$, then $j(x)=-140625/8$, $-9317$, $351/4$, $-162677523113838677$, or one of the 13 CM $j$-invariants in $\Q$.
\end{conjecture}
\begin{remark}
\filip{New remark.}
Let $x\in X_1(N)$ be a $\PP^1$-isolated point with $j(x) \in \Q$. One expects that the square-free part of the conductor of any $E$ with $j(E)=j(x)$ will be very small.  The reason for this is that any such isolated point will necessarily have small mod $p$ image for some $p$ dividing $N$. This in turn forces potentially good reduction on $E$ in all primes, except possibly for $p=2$, see \cite[Corollary 4.4]{Mazur1978} for one (out of many) example of this phenomenon. Indeed this is the case for all the curves appearing in \Cref{conj}; their conductors are either a square, or twice a square.

Hence one might except to find most or all of the $\PP^1$-isolated points among elliptic curves with relatively small conductor (and hence in the LMFDB). This observation perhaps makes the computational evidence supporting \Cref{conj} more convincing. 
\end{remark}

We say a point $x \in X_0(N)(\Q)$ is \textbf{exceptional} if $X_0(N)(\Q)$ is finite and $x$ corresponds to a non-CM elliptic curve over $\Q$. It is worth noting that the sporadic points on $X_1(21)$ associated to $j=-140625/8$ and the sporadic points on $X_1(37)$ associated to $j=-9317$ lie above exceptional points on $X_0(21)$ and $X_0(37)$, respectively. Thus it is natural to ask whether there exist other sporadic \abedit{$j$-invariants corresponding to} exceptional rational points on $X_0(N)$. Running Algorithm~\ref{alg:mainalgorithm} on the 14 $j$-invariants corresponding to exceptional rational points on $X_0(N)$ (see, for example, \cite[Table 4]{LRAnn}) shows that there are no additional sporadic $j$-invariants in this set.
\begin{theorem}
Let $E$ be an elliptic curve corresponding to an exceptional rational point on $X_0(N)$ for some positive integer $N$. \abedit{If $j(E)$ is sporadic, then $j(E)=-140625/8$ or $-9317$.}
\end{theorem}

@@

\subsection{Outline}
After providing relevant background material in Section \ref{sec:background}, we establish  preliminary mathematical results on genus 0 adelic images ($\S$\ref{sec:genus0}) and primitive points on modular curves ($\S$\ref{sec:PrimitivePoints}). We give an overview of the main algorithm in Section \ref{sec:mainalg}, with key sub-algorithms described in detail in Sections \ref{sec:madicrep} and \ref{sec:primitivedeg}. We address the validity of Algorithm \ref{alg:mainalgorithm} in Section \ref{sec:validity}. The output of the algorithm and its final analysis is discussed in Section \ref{sec:remainingfilters}.

@@

\subsection{Modular Curves}
For any subgroup $H \leq \GL_2(\Z/N\Z)$, we define the modular curve $X_H$ to be the coarse space of the stack $\mathcal{M}_H$, as defined in Deligne--Rapoport \cite{DR}. The curve $X_H$ is a scheme over $\text{Spec }\Z[1/N]$ and parametrizes generalized elliptic curves with $H$-level structure. \abedit{In particular, its $k$-rational points roughly classify elliptic curves over $k$ whose mod $N$ image is contained in $H$}; see, for example, \cite[$\S2.3$]{RSZB2022} for details. If we take
\[
B_1(N) =\left\{ \begin{pmatrix}
1 & * \\
0 & *
\end{pmatrix} \right \} \leq \GL_2(\Z/N\Z),
\]
then $X_{B_1(N)}=X_1(N)$, the modular curve whose noncuspidal points parametrize  elliptic curves with a distinguished point of order $N$. There is an analytic isomorphism between $X_1(N)(\mathbb{C})$ and the Riemann surface constructed as a quotient of the extended upper-half plane by the congruence subgroup
\[
\Gamma_1(N) \coloneqq\left\{ \begin{pmatrix}
a & b \\
c & d 
\end{pmatrix} \in \SL_2(\Z) \, : \, c \equiv 0 \pmod{N} \text{ and } a \equiv d \equiv 1 \pmod{N} \right \}, 
\]
with matrices acting via linear fractional transformations. If $N \geq 4$, then $\mathcal{M}_{B_1(N)}$ is its own coarse moduli space, and so noncuspidal $k$-rational points of $X_1(N)$ classify pairs $(E,P)$, where $E/k$ is an elliptic curve and $P\in E(k)$, up to $k$-isomorphism. 
%\abbey{I decided not to define the modular curve by its moduli problem as in \cite[$\S2$]{RouseDZB}, since the construction yields a stack in general... so it seems confusing to call it a modular *curve*. I've opted to go with the conventions of \cite[$\S2.3$]{RSZB2022}.}

@@

\subsection{Closed Points on Modular Curves}\abbey{I updated the first 2 paragraphs of this section.} To discuss isolated points on modular curves, we must consider closed points on $X_1(N)$, viewed always as a scheme over $\Q$. Let $k$ be a field with an embedding of $k$ into $\overline \Q$. Given an elliptic curve $E/k$ with $P \in E(k)$ of order $N$, the pair $(E,P)$ induces a $k$-valued point on $X_1(N)$ via the moduli interpretation described above. We denote this $k$-valued point by $(E,P)_k$, and by definition it corresponds to a morphism of $\Q$-schemes $f:\text{Spec } k \rightarrow X_1(N)$. The map $f$ sends the unique point of $\text{Spec } k$ to a point $x \in X$, and we call $x$ the \textbf{closed point} associated to $(E,P)$.\footnote{Note $x$ is indeed closed since $\Q(x)/\Q$ is finite; see, for example, \cite[Exercise 5.9, p.76] {Liu2002}.} We define the \textbf{degree} of $x$ to be the degree of the residue field $\Q(x)$ over $\Q$. Since there are many scheme-valued points which induce the same closed point, it is sometimes preferable to consider Galois orbits of points in $X_1(N)(\overline{\Q})$, which are in bijection with the set of closed points. Thus one could alternatively define the closed point associated to $(E,P)$ as the $\Gal_\Q$-orbit of $(E,P)_{\overline{\Q}}$.

@@

\begin{remark}\label{ResidueFieldRmk} In particular, if $E/\Q(j(E))$ does not have complex multiplication and $P=(x_0,y_0)$, then one can take $\Q(x)\cong\Q(j(E),x_0)$. See \cite[p. 107]{shimura}.
\end{remark}
\begin{example}
Let $E_1:y^2+xy+y=x^3-x^2-3x+3$ and $P_1=(-1,2)$ a point of order 7. Then $(E_1,P_1)$ gives a $\Q$-valued point on $X_1(7)/\Q$ and also a closed point $x\in X_1(7)$ of degree 1. On the other hand, let $E_2:y^2=x^3-43x-166$ and $P_2=(5,i\sqrt{256})$ a point of order 7. Then $(E_2,P_2)$ gives a $\Q(i)$-valued point on $X_1(7)/\Q$ and also a closed point $x\in X_1(7)$ of degree 1 by \abedit{Remark \ref{ResidueFieldRmk}}. In fact, both $(E_1,P_1)$ and $(E_2,P_2)$ induce the same geometric point on $X_1(7)$ since $(E_1,P_1)_{\overline{\Q}}=(E_2,P_2)_{\overline{\Q}}$.

@@

To determine its degree, let $\Gamma$ and $\Gamma'$ be the intersection with $\SL_2(\Z)$ of the inverse image of $G$ and $G'$ in $\GL_2(\Z)$. Over $\bC$, the morphism $f$ is the quotient map $\Gamma \backslash \mathcal{H} \rightarrow \Gamma' \backslash \mathcal{H}$. Since the kernel of the action of $\SL_2(\Z)$ on $\mathcal H$ is $\pm I$, the degree of $f$ is as claimed.
    \end{proof}
Let $a$ and $b$ be positive integers. Taking $G=\pi^{-1}(B_1(ab))$ and $G'=\pi^{-1}(B_1(a))$ gives the following corollary, which under the moduli interpretation corresponds to sending $(E,P)$ to $(E,bP)$.    \begin{corollary}\label{Cor:DegreeFormula}
                For positive integers $a$ and $b$, the natural $\Q$-rational map $f\colon X_1(ab) \rightarrow X_1(a)$ has
                \[
                    \deg(f)=
                    c_{f}\cdot b^2 \prod_{p \mid b,\, p \nmid a}
                    \left(1-\frac{1}{p^2}\right),
                \]
                where $c_{f}=1/2$ if $a \leq 2$ and $ab>2$ and $c_{f}=1$ otherwise. 
    \end{corollary}

\subsection{Galois Representations} If $E$ is an elliptic curve defined over a number field $k$, then $\Gal_k$ acts naturally on the points of $E$. On torsion points, this action is described by the \textbf{adelic Galois representation} associated to $E/k$, 
\[
\rho_{E}: \Gal_k \rightarrow \Aut(E(\overline{k})_{\text{tors}}) \cong \GL_2(\widehat{\Z}). 
\]
From this we can obtain two other Galois representations. On the one hand, fixing a positive integer $m$, we can choose to record the action of $\Gal_k$ on points \abedit{whose order is divisible only by those primes dividing $m$}. This is the \textbf{$m$-adic Galois representation} associated to $E$: 
\[
\rho_{E,m^{\infty}}\colon \Gal_k \xrightarrow{\rho_E} \GL_2(\widehat{\Z}) \cong  \prod_{p \text{ prime}} \GL_2(\Z_{p}) \xrightarrow{\proj} \prod_{p \mid m} \GL_2(\Z_{p}).
\]
In particular, if $m=\ell$ is a prime number, we recover the standard $\ell$-adic representation associated to $E$. Alternatively, we may wish to record Galois action only on points of order dividing $m$; we use $E[m]$ to denote the finite subgroup of such points. This gives the \textbf{mod $m$ Galois representation} associated to $E$,
\[
\rho_{E,m}\colon \Gal_k \rightarrow \Aut(E[m]) \cong \GL_2(\Z/m\Z).
\]
Note that $\rho_{E,m}$ agrees with the reduction of $\rho_E$ mod $m$.
